---
title: "IS5 in R: Stats Starts Here (Chapter 5)"
author: "Margaret Chien and Nicholas Horton (nhorton@amherst.edu)"
date: "June 16, 2018"
output: 
  pdf_document:
    fig_height: 4
    fig_width: 6
  html_document:
    fig_height: 3
    fig_width: 5
  word_document:
    fig_height: 4
    fig_width: 6
---


```{r, include = FALSE}
# Don't delete this chunk if you are using the mosaic package
# This loads the mosaic and dplyr packages
library(mosaic)
library(readr)
library(janitor)
```

```{r, include = FALSE}
# knitr settings to control how R chunks work.
require(knitr)
opts_chunk$set(
  tidy = FALSE,     # display code as typed
  size = "small"    # slightly smaller font for code
)
```

## Introduction and background 

This document is intended to help describe how to undertake analyses introduced 
as examples in the Fifth Edition of *Intro Stats* (2018) by De Veaux, Velleman, and Bock.
More information about the book can be found at http://wps.aw.com/aw_deveaux_stats_series.  This
file as well as the associated R Markdown reproducible analysis source file used to create it can be found at http://nhorton.people.amherst.edu/is5.

This work leverages initiatives undertaken by Project MOSAIC (http://www.mosaic-web.org), an NSF-funded effort to improve the teaching of statistics, calculus, science and computing in the undergraduate curriculum. In particular, we utilize the `mosaic` package, which was written to simplify the use of R for introductory statistics courses. A short summary of the R needed to teach introductory statistics can be found in the mosaic package vignettes (http://cran.r-project.org/web/packages/mosaic).  A paper describing the mosaic approach was published in t
he *R Journal*: https://journal.r-project.org/archive/2017/RJ-2017-024.

## Chapter 5: The Standard Deviation as a Ruler and the Normal Model

```{r}
library(mosaic)
library(readr)
library(janitor)
WomenHeptathlon2016 <- read_csv("http://nhorton.people.amherst.edu/is5/data/Womens_Heptathlon_2016.csv") %>%
  clean_names()
favstats(~ long_jump, data = WomenHeptathlon2016)
favstats(~ x200m, data = WomenHeptathlon2016)
with(WomenHeptathlon2016, stem(x200m))
with(WomenHeptathlon2016, stem(long_jump))
```

### Section 5.1: Using the Standard Deviation to Standardize Values

```{r}
filter(WomenHeptathlon2016, last_name == "Thiam")
# calculate z-score with mean and sd from favstats
(6.58 - 6.17)/.247 # long jump
filter(WomenHeptathlon2016, last_name == "Johnson-Thompson")
```

### Section 5.2: Shifting and Scaling

#### Shifting to Adjust the Center

```{r}
MenWeight <- read_csv("http://nhorton.people.amherst.edu/is5/data/Mens_Weights.csv") %>%
  clean_names()
gf_histogram(~ weight_in_kg, data = MenWeight, xlab = "Weight (kg)", ylab = "# of Men", binwidth = 10)
gf_boxplot(weight_in_kg ~ 1, data = MenWeight, xlab = "", ylab = "Weight (kg)") %>%
  gf_refine(coord_flip())
```

```{r}
favstats(~ weight_in_kg, data = MenWeight)
gf_histogram(~ (weight_in_kg - 74), data = MenWeight, xlab = "Kg Above Recommended Weight", ylab = "# of Men", binwidth = 10)
```

#### Rescaling to Adjust the Scale

```{r}
gf_histogram(~ weight_in_kg, data = MenWeight, xlab = "Kilograms", ylab = "# of Men", binwidth = 10)
gf_histogram(~ weight_in_pounds, data = MenWeight, xlab = "Pounds", ylab = "# of Men", binwidth = 20)
favstats(~ weight_in_kg, data = MenWeight)
favstats(~ weight_in_pounds, data = MenWeight)
library(tidyr) # for gather() function
MenWeight %>%
  gather(key = weighttype, value = weight, weight_in_kg, weight_in_pounds) %>%
  gf_boxplot(weight ~ weighttype, ylab = "", xlab = "Weight Type")
```

#### Shifting, Scaling, and the *z*-Scores

### Section 5.3: Normal Models 

#### The 68-95-99.7 Rule

See display on page 129.  

#### Example 5.4: Using the 68-95-99.7 Rule



### Section 5.4: Working with Normal Percentiles

### Section 5.5: Normal Probability Plots