---
title: "IS5 in R: Understanding and Comparing Distributions (Chapter 4)"
author: "Margaret Chien and Nicholas Horton (nhorton@amherst.edu)"
date: "July 12, 2018"
output: 
  pdf_document:
    fig_height: 3
    fig_width: 5
  html_document:
    fig_height: 3
    fig_width: 5
  word_document:
    fig_height: 4
    fig_width: 6
---


```{r, include = FALSE}
# Don't delete this chunk if you are using the mosaic package
# This loads the mosaic and dplyr packages
library(mosaic)
library(readr)
library(janitor)
```

```{r, include = FALSE}
# knitr settings to control how R chunks work.
require(knitr)
opts_chunk$set(
  tidy = FALSE,     # display code as typed
  size = "small"    # slightly smaller font for code
)
```

## Introduction and background 

This document is intended to help describe how to undertake analyses introduced 
as examples in the Fifth Edition of *Intro Stats* (2018) by De Veaux, Velleman, and Bock.
More information about the book can be found at http://wps.aw.com/aw_deveaux_stats_series.  This
file as well as the associated R Markdown reproducible analysis source file used to create it can be found at http://nhorton.people.amherst.edu/is5.

This work leverages initiatives undertaken by Project MOSAIC (http://www.mosaic-web.org), an NSF-funded effort to improve the teaching of statistics, calculus, science and computing in the undergraduate curriculum. In particular, we utilize the `mosaic` package, which was written to simplify the use of R for introductory statistics courses. A short summary of the R needed to teach introductory statistics can be found in the mosaic package vignettes (http://cran.r-project.org/web/packages/mosaic).  A paper describing the mosaic approach was published in the *R Journal*: https://journal.r-project.org/archive/2017/RJ-2017-024.

## Chapter 4: Understanding and Comparing Distributions

```{r}
library(mosaic)
library(readr)
library(janitor)
HopkinsForest <- read_csv("http://nhorton.people.amherst.edu/is5/data/Hopkins_Forest.csv") %>%
  clean_names()
names(HopkinsForest)
```

By default, `read_csv()` prints the variable names. These messages can be suppressed using the `message = FALSE` code chunk option to save space and improve readability.   
Here we use the `clean_names()` function from the `janitor` package to sanitize the names of the columns (which would otherwise contain special characters or whitespace). You can use the `names()` function to check the cleaned names.

```{r}
gf_histogram(~ avg_wind_mph, data = HopkinsForest, binwidth = 0.5, center = 0.25)
favstats(~ avg_wind_mph, data = HopkinsForest)
```

### Section 4.1: Displays for Comparing Groups

#### Histograms

```{r}
HopkinsForest <- HopkinsForest %>%
  mutate(catmonth = ifelse(month <= 9 & month >= 4, "Spring/Summer", "Fall/Winter"))
gf_histogram(~ avg_wind_mph, data = HopkinsForest, binwidth = 0.5, center = 0.25) %>%
  gf_facet_wrap(~ catmonth)
favstats(~ avg_wind_mph | catmonth, data = HopkinsForest)
```

#### Example 4.1: Comparing Groups with Stem-And-Leaf

```{r}
NestEgg <- read_csv("http://nhorton.people.amherst.edu/is5/data/Nest_Egg_Index.csv") %>%
  clean_names()
with(NestEgg, stem(nest_egg_index))
```

#### Boxplots

```{r}
gf_boxplot(~ avg_wind_mph, data = HopkinsForest)  # or gf_boxplot(X ~ 1)
```

As noted in the book, boxplots are most useful to compare distributions.  Here we replicate the single boxplot 
from page 98 but don't recommend the use of single boxplots.

```{r}
gf_boxplot(avg_wind_mph ~ as.factor(month), data = HopkinsForest) %>%
  gf_labs(x = "month", y = "average wind speed (mph)")
```

Here we replicate the results from Figure 4.3 (page 99).

We use `as.factor()` to convert a variable into a factor.  
We also use `gf_labs()` to clean up the code for the first line and improve readability.  

#### Example 4.2: Comparing Groups with Boxplots

```{r}
Coasters <- read_csv("http://nhorton.people.amherst.edu/is5/data/Coasters_2015.csv")
gf_boxplot(Speed ~ Track, data = Coasters)
```

#### Step-By-Step Example: Comparing Groups

```{r}
Cups <- read_csv("http://nhorton.people.amherst.edu/is5/data/Cups.csv")
favstats(~ Difference | Container, data = Cups)
gf_boxplot(Difference ~ Container, data = Cups) 
```

#### Just Checking

```{r, warning=FALSE}
Flights <- read_csv("http://nhorton.people.amherst.edu/is5/data/Flights_on_time_2016.csv") %>%
  clean_names()
gf_histogram(~ ontime_pct, data = Flights, binwidth = 2, center = 1) %>%
  gf_labs(x = "Ontime %", y = "$ of Months")
gf_boxplot(~ ontime_pct, data = Flights)
gf_boxplot(ontime_pct ~ month, data = Flights)
```

#### Random Matters

```{r}
CarSpeeds <- read_csv("http://nhorton.people.amherst.edu/is5/data/Car_speeds.csv")
gf_boxplot(speed ~ direction, data = CarSpeeds)
```

See histogram on page 103.  

### Section 4.2: Outliers

```{r}
gf_boxplot(avg_wind_mph ~ as.factor(month), data = HopkinsForest) %>%
  gf_labs(x = "Month", y = "Average Wind Speed (mph)")
```

#### Example 4.3: Checking Out the Outliers

```{r}
gf_boxplot(Speed ~ Track, data = Coasters)
```

### Section 4.3: Re-Expressing Data: A First Look

#### Re-Expressing to Improve Symmetry

```{r}
CEOComp <- read_csv("http://nhorton.people.amherst.edu/is5/data/CEO_Compensation_2014.csv") %>%
  clean_names()
# Figure 4.6, page 105
gf_histogram(~ ceo_compensation_m, data = CEOComp, binwidth = 2.5, center = 2.5/2) %>%
  gf_labs(x = "Compensation (M$)", y = "Millions of $")
gf_boxplot(~ ceo_compensation_m, data = CEOComp) %>%
  gf_labs(x = "Compensation (M$)", y = "Millions of $")

# Figure 4.7, page 106
gf_histogram(~ log(ceo_compensation_m), data = CEOComp, binwidth = .224, center = .112) %>%
  gf_labs(x = "Log (compensation)", y = "# of CEOs")
```

#### Re-Expression to Equalize Spread Across Groups

```{r}
PassiveSmoke <- read_csv("http://nhorton.people.amherst.edu/is5/data/Passive_smoke.csv")
gf_boxplot(cotinine ~ smoke_exposure, data = PassiveSmoke) %>%
  gf_labs(x = "Smoke Exposure", y = "Cotinine (ng/ml)")
gf_boxplot(log(cotinine) ~ smoke_exposure, data = PassiveSmoke) %>%
  gf_labs(x = "Smoke Exposure", y = "Log(cotinine)")
```
