---
title: "IS5 in R: Displaying and Describing Data (Chapter 2)"
author: "Margaret Chien and Nicholas Horton (nhorton@amherst.edu)"
date: "June 10, 2018"
output: 
  pdf_document:
    fig_height: 4
    fig_width: 6
  html_document:
    fig_height: 3
    fig_width: 5
  word_document:
    fig_height: 4
    fig_width: 6
---


```{r, include = FALSE}
# Don't delete this chunk if you are using the mosaic package
# This loads the mosaic and dplyr packages
require(mosaic)
```

```{r, include = FALSE}
# knitr settings to control how R chunks work.
require(knitr)
opts_chunk$set(
  tidy = FALSE,     # display code as typed
  size = "small"    # slightly smaller font for code
)
```

## Introduction and background 

This document is intended to help describe how to undertake analyses introduced 
as examples in the Fifth Edition of *Intro Stats* (2018) by De Veaux, Velleman, and Bock.
More information about the book can be found at http://wps.aw.com/aw_deveaux_stats_series.  This
file as well as the associated R Markdown reproducible analysis source file used to create it can be found at http://nhorton.people.amherst.edu/is5.

This work leverages initiatives undertaken by Project MOSAIC (http://www.mosaic-web.org), an NSF-funded effort to improve the teaching of statistics, calculus, science and computing in the undergraduate curriculum. In particular, we utilize the `mosaic` package, which was written to simplify the use of R for introductory statistics courses. A short summary of the R needed to teach introductory statistics can be found in the mosaic package vignettes (http://cran.r-project.org/web/packages/mosaic).  A paper describing the mosaic approach was published in the *R Journal*: https://journal.r-project.org/archive/2017/RJ-2017-024.

## Chapter 2: Displaying and Describing Data

### Section 2.1: Summarizing and Displaying a Categorical Variable

See displays on page 19-23.

```{r message = FALSE}
library(readr)
library(janitor) #for column names
options(digits = 3)
Titanic <- read_csv("http://nhorton.people.amherst.edu/is5/data/Titanic.csv")
tally(~ Class, data = Titanic)
tally(~ Class, format = "percent", data = Titanic)
gf_bar(~ Class, data = Titanic)
```

### Section 2.2: Displaying a Quantitative Variable

#### Ages of Those Aboard the the Titanic

```{r}
gf_histogram(~ Age, data = Titanic, bins = 15, ylab = "# of People")
```

#### Earthquakes and Tsunamis

```{r}
Earthquakes <- read_csv("http://nhorton.people.amherst.edu/is5/data/Tsunamis_2016.csv")
gf_histogram(~ Primary_Magnitude, data = Earthquakes, bins = 12, 
             ylab = "Earthquakes", xlab = "Magnitude")
```

#### Stem-and-Leaf Displays

See page 26.

```{r}
Pulse_rates <- read_csv("http://nhorton.people.amherst.edu/is5/data/Pulse_rates.csv")
gf_histogram(~ Pulse, data = Pulse_rates, bins = 7)
with(Pulse_rates, stem(Pulse))
```

#### Dotplot

```{r}
Derby <- read_csv("http://nhorton.people.amherst.edu/is5/data/Kentucky_Derby_2016.csv")
gf_dotplot(~ Time_Sec, data = Derby, binwidth = 1)
```

#### Density Plots

```{r}
gf_dens(~ Age, data = Titanic, ylab = "Density")
```

### Section 2.3: Shape

See displays on pages 28-29.

#### Consumer Price Index

```{r}
CPI <- read_csv("http://nhorton.people.amherst.edu/is5/data/CPI_Worldwide.csv") %>%
  clean_names()
gf_histogram(~ consumer_price_index, data = CPI, ylab = "# of Cities", 
             xlab = "Consumer Price Index")
```

We can use `clean_names()` from the `janitor` package to format the names of the columns when necessary.  

#### Credit Card Expenditures

```{r}
CreditCardEx <- read_csv("http://nhorton.people.amherst.edu/is5/data/Credit_card_charges.csv") %>%
  clean_names()
gf_histogram(~ charges, data = CreditCardEx, ylab = "# of Customers", 
             xlab = "Average Monthly Expenditure ($)")
```

### Section 2.4: Center

#### Finding Median and Mean

```{r}
TitanicCrew <- filter(Titanic, Class == "Crew")
gf_histogram(~ Age, data = TitanicCrew)
favstats(~ Age, data = TitanicCrew)
```

See displays on pages 32 and 33.

### Section 2.5: Spread

#### The Range

```{r}
range(~ Age, data = TitanicCrew)
diff(range(~ Age, data = TitanicCrew))
```

#### The Interquartile Range

```{r}
favstats(~ Age, data = TitanicCrew)
IQR(~ Age, data = TitanicCrew)
```

We can find IQR by subtracting Q1 from Q3.  

#### Standard Deviation  

```{r}
sd(~ Age, data = TitanicCrew)
var(~ Age, data = TitanicCrew)
```

### Summarizing a Distribution

```{r}
Nissan <- read_csv("http://nhorton.people.amherst.edu/is5/data/Nissan.csv")
gf_histogram(~ mpg, data = Nissan, bins = 18, xlab = "Fuel Efficiency (mpg)", 
             ylab = "# of Fill-ups")
favstats(~ mpg, data = Nissan)
```

### Random Matters

```{r}
Commute <- read_csv("http://nhorton.people.amherst.edu/is5/data/Population_Commute_Times.csv") %>%
  clean_names()
gf_histogram(~ commute_time, data = Commute, binwidth = 10, xlab = "Commute Time (min)", 
             ylab = "# of Employees")
```
