---
title: "IS5 in R: Linear Regression (Chapter 7)"
author: "Margaret Chien and Nicholas Horton (nhorton@amherst.edu)"
date: "June 16, 2018"
output: 
  pdf_document:
    fig_height: 4
    fig_width: 6
  html_document:
    fig_height: 3
    fig_width: 5
  word_document:
    fig_height: 4
    fig_width: 6
---


```{r, include = FALSE}
# Don't delete this chunk if you are using the mosaic package
# This loads the mosaic and dplyr packages
library(mosaic)
library(readr)
library(janitor)
```

```{r, include = FALSE}
# knitr settings to control how R chunks work.
require(knitr)
opts_chunk$set(
  tidy = FALSE,     # display code as typed
  size = "small"    # slightly smaller font for code
)
```

## Introduction and background 

This document is intended to help describe how to undertake analyses introduced 
as examples in the Fifth Edition of *Intro Stats* (2018) by De Veaux, Velleman, and Bock.
More information about the book can be found at http://wps.aw.com/aw_deveaux_stats_series.  This
file as well as the associated R Markdown reproducible analysis source file used to create it can be found at http://nhorton.people.amherst.edu/is5.

This work leverages initiatives undertaken by Project MOSAIC (http://www.mosaic-web.org), an NSF-funded effort to improve the teaching of statistics, calculus, science and computing in the undergraduate curriculum. In particular, we utilize the `mosaic` package, which was written to simplify the use of R for introductory statistics courses. A short summary of the R needed to teach introductory statistics can be found in the mosaic package vignettes (http://cran.r-project.org/web/packages/mosaic).  A paper describing the mosaic approach was published in the *R Journal*: https://journal.r-project.org/archive/2017/RJ-2017-024.

## Chapter 7: Linear Regression

```{r}
library(mosaic)
library(readr)
library(janitor)
BurgerKing <- read_csv("http://nhorton.people.amherst.edu/is5/data/Burger_King_items.csv") %>%
  clean_names()
gf_point(fat_g ~ protein_g, data = BurgerKing, xlab = "Protein (g)", ylab = "Fat (g)")
```

### Section 7.1: Least Squares: The Line of "Best Fit"

See display on page 197.

We can calculate the residual for a particular value with 31 grams of protein.  

```{r}
burgerlm <- lm(fat_g ~ protein_g, data = BurgerKing)
burgerfun <- makeFun(burgerlm)
burgerfun(31)
```

### Section 7.2: The Linear Model

```{r}
coef(burgerlm)
burgerfun(0)
burgerfun(32) - burgerfun(31)
```

#### Example 7.1: A Linear Model for Hurricanes

```{r}
Hurricanes <- read_csv("http://nhorton.people.amherst.edu/is5/data/Hurricanes_2015.csv") %>%
  clean_names()
gf_point(max_wind_speed_kts ~ central_pressure_mb, data = Hurricanes, xlab = "Central Pressure (mb)", ylab = "Max Wind Speed (kts)") %>%
  gf_lm()
```

### Section 7.3: Finding the Least Squares Line

#### Example 7.2: Finding the Regression Equation

```{r}
favstats(~ protein_g, data = BurgerKing)
favstats(~ fat_g, data = BurgerKing)
sx <- sd(~ protein_g, data = BurgerKing)
sx
sy <- sd(~ fat_g, data = BurgerKing)
sy
r <- cor(protein_g ~ fat_g, data = BurgerKing)
r   # same as cor(fat_g ~ protein_g)!
r*sy/sx
coef(burgerlm)[2]
```

#### Step-by-Step Example: Calculating a Regression Equation

```{r}
TompkinsBridges <- read_csv("http://nhorton.people.amherst.edu/is5/data/Tompkins_county_bridges_2016.csv") %>%
  clean_names()
gf_point(condition ~ age_at_inspection, data = TompkinsBridges, xlab = "Age at Inspection", ylab = "Condition")
```

See calculations on page 203.

### Section 7.4: Regression to the Mean

See display on page 205.

### Section 7.5: Examining the Residuals

```{r}
msummary(burgerlm)
gf_point(resid(burgerlm) ~ protein_g, data = BurgerKing) %>%
  gf_lm() %>%
  gf_labs(x = "Protein (g)", y = "Residuals (g fat)")
gf_histogram(~ resid(burgerlm), bins = 8) %>%
  gf_labs(x = "Residuals (g fat)", y = "# of Residuals")
```

### Section 7.6: $R^2$--The Variation Accounted For by the Model

```{r}
rsquared(burgerlm)
```

### Section 7.7: Regression Assumptions and Conditions