---
title: "IS5 in R: More About Tests and Intervals (Chapter 16)"
author: "Margaret Chien and Nicholas Horton (nhorton@amherst.edu)"
date: "July 12, 2018"
output: 
  pdf_document:
    fig_height: 3
    fig_width: 5
  html_document:
    fig_height: 3
    fig_width: 5
  word_document:
    fig_height: 4
    fig_width: 6
---


```{r, include = FALSE}
# Don't delete this chunk if you are using the mosaic package
# This loads the mosaic and dplyr packages
library(mosaic)
```

```{r, include = FALSE}
# knitr settings to control how R chunks work.
require(knitr)
opts_chunk$set(
  tidy = FALSE,     # display code as typed
  size = "small"    # slightly smaller font for code
)
```

## Introduction and background 

This document is intended to help describe how to undertake analyses introduced 
as examples in the Fifth Edition of *Intro Stats* (2018) by De Veaux, Velleman, and Bock.
More information about the book can be found at http://wps.aw.com/aw_deveaux_stats_series.  This
file as well as the associated R Markdown reproducible analysis source file used to create it can be found at http://nhorton.people.amherst.edu/is5.

This work leverages initiatives undertaken by Project MOSAIC (http://www.mosaic-web.org), an NSF-funded effort to improve the teaching of statistics, calculus, science and computing in the undergraduate curriculum. In particular, we utilize the `mosaic` package, which was written to simplify the use of R for introductory statistics courses. A short summary of the R needed to teach introductory statistics can be found in the mosaic package vignettes (http://cran.r-project.org/web/packages/mosaic).  A paper describing the mosaic approach was published in the *R Journal*: https://journal.r-project.org/archive/2017/RJ-2017-024.

## Chapter 16: More About Tests and Intervals

```{r}
library(mosaic)
```

### Section 16.1: Interpreting P-Values

#### What to Do with a Low P-Value

#### What to Do with a High P-Value

```{r, fig.width=7}
# curve on page 511
xqnorm(p = .467, mean = 0, sd = 1, verbose = FALSE)
```

### Section 16.2: Alpha Levels and Critical Values

```{r}
# Figure 16.1, page 513
xpnorm(q = c(-1.96, 1.96), mean = 0, sd = 1, verbose = FALSE)
```

### Section 16.3: Practical vs. Statistical Significance

### Section 16.4: Errors

#### Power

#### Effect Size

#### A Picture Worth $\frac{1}{P(z > 3.09)}$ Words

```{r}
# Figure 16.2, page 520
xpnorm(q = 1.96, mean = 0, sd = 1, verbose = FALSE)
# Not sure how to do this

#ggplot(data.frame(x = c(-4, 8)), aes(x)) + 
#  stat_function(fun = dnorm, args = list(mean = 0, sd = 1), col='red') +
#  stat_function(fun = dnorm, args = list(mean = 4, sd = 1), col='blue') +
#  geom_vline()
```