---
title: "IS5 in R: Sample Surveys (Chapter 10)"
author: "Margaret Chien and Nicholas Horton (nhorton@amherst.edu)"
date: "July 10, 2018"
output: 
  pdf_document:
    fig_height: 4
    fig_width: 6
  html_document:
    fig_height: 3
    fig_width: 5
  word_document:
    fig_height: 4
    fig_width: 6
---


```{r, include = FALSE}
# Don't delete this chunk if you are using the mosaic package
# This loads the mosaic and dplyr packages
library(mosaic)
library(readr)
library(janitor)
```

```{r, include = FALSE}
# knitr settings to control how R chunks work.
require(knitr)
opts_chunk$set(
  tidy = FALSE,     # display code as typed
  size = "small"    # slightly smaller font for code
)
```

## Introduction and background 

This document is intended to help describe how to undertake analyses introduced 
as examples in the Fifth Edition of *Intro Stats* (2018) by De Veaux, Velleman, and Bock.
More information about the book can be found at http://wps.aw.com/aw_deveaux_stats_series.  This
file as well as the associated R Markdown reproducible analysis source file used to create it can be found at http://nhorton.people.amherst.edu/is5.

This work leverages initiatives undertaken by Project MOSAIC (http://www.mosaic-web.org), an NSF-funded effort to improve the teaching of statistics, calculus, science and computing in the undergraduate curriculum. In particular, we utilize the `mosaic` package, which was written to simplify the use of R for introductory statistics courses. A short summary of the R needed to teach introductory statistics can be found in the mosaic package vignettes (http://cran.r-project.org/web/packages/mosaic).  A paper describing the mosaic approach was published in the *R Journal*: https://journal.r-project.org/archive/2017/RJ-2017-024.

## Chapter 10: Sample Surveys

```{r}
library(mosaic)
library(readr)
library(janitor)
```

### Section 10.1: The Three Big Ideas of Sampling

### Section 10.2: Populations and Parameters

### Section 10.3: Simple Random Samples

#### Random Matters

```{r}
Births <- read_csv("http://nhorton.people.amherst.edu/is5/data/AllBirths1998.csv")
```

By default, `read_csv()` prints the variable names. These messages can be suppressed using the `message = FALSE` code chunk option to save space and improve readability.  

```{r}
gf_histogram(~ birthweight, data = Births, binwidth = 125, center = 62.5) %>%
  gf_labs(x = "Weight (g)", y = "Births")
```

```{r}
# Samples of 100
set.seed(12452)
gf_histogram(~ birthweight, data = sample(Births, size = 100), binwidth = 200, center = 100) %>%
  gf_labs(x = "Weight 1", y = "")
gf_histogram(~ birthweight, data = sample(Births, size = 100), binwidth = 200, center = 100) %>%
  gf_labs(x = "Weight 2", y = "")
gf_histogram(~ birthweight, data = sample(Births, size = 100), binwidth = 200, center = 100) %>%
  gf_labs(x = "Weight 3", y = "")
gf_histogram(~ birthweight, data = sample(Births, size = 100), binwidth = 200, center = 100) %>%
  gf_labs(x = "Weight 4", y = "")
gf_histogram(~ birthweight, data = sample(Births, size = 100), binwidth = 200, center = 100) %>%
  gf_labs(x = "Weight 5", y = "")
```

```{r}
# Samples of 250
set.seed(12452)
gf_histogram(~ birthweight, data = sample(Births, size = 250), binwidth = 125, center = 62.5) %>%
  gf_labs(x = "Weight 1", y = "")
gf_histogram(~ birthweight, data = sample(Births, size = 250), binwidth = 125, center = 62.5) %>%
  gf_labs(x = "Weight 2", y = "")
gf_histogram(~ birthweight, data = sample(Births, size = 250), binwidth = 125, center = 62.5) %>%
  gf_labs(x = "Weight 3", y = "")
gf_histogram(~ birthweight, data = sample(Births, size = 250), binwidth = 125, center = 62.5) %>%
  gf_labs(x = "Weight 4", y = "")
gf_histogram(~ birthweight, data = sample(Births, size = 100), binwidth = 125, center = 62.5) %>%
  gf_labs(x = "Weight 5", y = "")
```

```{r}
# Samples of 1000
set.seed(12452)
gf_histogram(~ birthweight, data = sample(Births, size = 1000), binwidth = 200, center = 100) %>%
  gf_labs(x = "Weight 1", y = "")
gf_histogram(~ birthweight, data = sample(Births, size = 1000), binwidth = 200, center = 100) %>%
  gf_labs(x = "Weight 2", y = "")
gf_histogram(~ birthweight, data = sample(Births, size = 1000), binwidth = 200, center = 100) %>%
  gf_labs(x = "Weight 3", y = "")
gf_histogram(~ birthweight, data = sample(Births, size = 1000), binwidth = 200, center = 100) %>%
  gf_labs(x = "Weight 4", y = "")
gf_histogram(~ birthweight, data = sample(Births, size = 1000), binwidth = 200, center = 100) %>%
  gf_labs(x = "Weight 5", y = "")
```

### Section 10.4: Other Sampling Designs

### Section 10.5: From the Population to the Sample: You Can't Always Get What You Want

### Section 10.6: The Valid Survey

### Section 10.7: Common Sampling Mistakes, or How to Sample Badly