---
title: "IS5 in R: Sampling Distribution Models and Confidence Intervals for Proportions (Chapter 13)"
author: "Margaret Chien and Nicholas Horton (nhorton@amherst.edu)"
date: "July 10, 2018"
output: 
  pdf_document:
    fig_height: 4
    fig_width: 6
  html_document:
    fig_height: 3
    fig_width: 5
  word_document:
    fig_height: 4
    fig_width: 6
---


```{r, include = FALSE}
# Don't delete this chunk if you are using the mosaic package
# This loads the mosaic and dplyr packages
library(mosaic)
library(readr)
library(janitor)
```

```{r, include = FALSE}
# knitr settings to control how R chunks work.
require(knitr)
opts_chunk$set(
  tidy = FALSE,     # display code as typed
  size = "small"    # slightly smaller font for code
)
```

## Introduction and background 

This document is intended to help describe how to undertake analyses introduced 
as examples in the Fifth Edition of *Intro Stats* (2018) by De Veaux, Velleman, and Bock.
More information about the book can be found at http://wps.aw.com/aw_deveaux_stats_series.  This
file as well as the associated R Markdown reproducible analysis source file used to create it can be found at http://nhorton.people.amherst.edu/is5.

This work leverages initiatives undertaken by Project MOSAIC (http://www.mosaic-web.org), an NSF-funded effort to improve the teaching of statistics, calculus, science and computing in the undergraduate curriculum. In particular, we utilize the `mosaic` package, which was written to simplify the use of R for introductory statistics courses. A short summary of the R needed to teach introductory statistics can be found in the mosaic package vignettes (http://cran.r-project.org/web/packages/mosaic).  A paper describing the mosaic approach was published in the *R Journal*: https://journal.r-project.org/archive/2017/RJ-2017-024.

## Chapter 13: Sampling Distribution Models and Confidence Intervals for Proportions

```{r}
library(mosaic)
library(readr)
library(janitor)
Babies <- read_csv("http://nhorton.people.amherst.edu/is5/data/Babysamp_98.csv") %>%
  clean_names()
```

By default, `read_csv()` prints the variable names. These messages can be suppressed using the `message = FALSE` code chunk option to save space and improve readability.  
Here we use the `clean_names()` function from the `janitor` package to sanitize the names of the columns (which would otherwise contain special characters or whitespace).   

```{r}
gf_histogram(~ gestation, binwidth = 1, center = .5, fill = ~ preemie, data = Babies) %>%
  gf_labs(x = "Gestation Time (weeks)", y = "Births")
```

### Section 13.1: The Sampling Distribution Model for a Proportion

#### The Normal Model

### Section 13.2: When Does the Normal Model Work? Assumptions and Conditions

#### Random Matters: Does the Normal Model Always Work? Sampling Distributions for Other Statistics

```{r}
BodyFat <- read_csv("http://nhorton.people.amherst.edu/is5/data/Bodyfat.csv") %>%
  clean_names()
set.seed(3245) # For reproducibility
numsim <- 1000  # Number of samples

# What does do() do?
favstats(~ weight, data = sample(BodyFat, 10)) # favstats of one random sample of 10
favstats(~ weight, data = sample(BodyFat, 10)) # favstats of another random sample

do(2) * favstats(~ weight, data = sample(BodyFat, 10)) # finds favstats twice

# For the visualization, we need 1,000 favstats
bodyfatsamples <- do(numsim) * favstats(~ weight, data = sample(BodyFat, 10))
```

Here, the `do()` function finds, 1,000 times, the `favstats()` of a random sample of 10 `BodyFat` weights. 

```{r}
bodyfatsamples <- bodyfatsamples %>%
  clean_names()
names(bodyfatsamples)
gf_histogram(~ median, data = bodyfatsamples, binwidth = 3, center = 1.5) %>%
  gf_labs(x = "Medians", y = "# of Samples")
gf_histogram(~ sd^2, data = bodyfatsamples) %>%
  gf_labs(x = "Variance", y = "# of Samples")
gf_histogram(~ min, data = bodyfatsamples, binwidth = 3, center = 1.5) %>%
  gf_labs(x = "Minimums", y = "# of Samples")
```

### Section 13.3: A Confidence Interval for a Proportion

### Section 13.4: Interpreting Confidence Intervals: What Does 95% Confidence Really Mean?

```{r}
do(20) * confint(~ preemie, data = sample(Babies, size = 100))
```

### Section 13.5: Margin of Error: Certainty vs. Precision

### Section 13.6: Choosing the Sample Size