---
title: "IS5 in R: Paired Samples and Blocks (Chapter 18)"
author: "Margaret Chien and Nicholas Horton (nhorton@amherst.edu)"
date: "July 11, 2018"
output: 
  pdf_document:
    fig_height: 3
    fig_width: 5
  html_document:
    fig_height: 3
    fig_width: 5
  word_document:
    fig_height: 4
    fig_width: 6
---


```{r, include = FALSE}
# Don't delete this chunk if you are using the mosaic package
# This loads the mosaic and dplyr packages
library(mosaic)
library(readr)
library(janitor)
library(tidyr)
```

```{r, include = FALSE}
# knitr settings to control how R chunks work.
require(knitr)
opts_chunk$set(
  tidy = FALSE,     # display code as typed
  size = "small"    # slightly smaller font for code
)
```

## Introduction and background 

This document is intended to help describe how to undertake analyses introduced 
as examples in the Fifth Edition of *Intro Stats* (2018) by De Veaux, Velleman, and Bock.
More information about the book can be found at http://wps.aw.com/aw_deveaux_stats_series.  This
file as well as the associated R Markdown reproducible analysis source file used to create it can be found at http://nhorton.people.amherst.edu/is5.

This work leverages initiatives undertaken by Project MOSAIC (http://www.mosaic-web.org), an NSF-funded effort to improve the teaching of statistics, calculus, science and computing in the undergraduate curriculum. In particular, we utilize the `mosaic` package, which was written to simplify the use of R for introductory statistics courses. A short summary of the R needed to teach introductory statistics can be found in the mosaic package vignettes (http://cran.r-project.org/web/packages/mosaic).  A paper describing the mosaic approach was published in the *R Journal*: https://journal.r-project.org/archive/2017/RJ-2017-024.

## Chapter 18: Paired Samples and Blocks

```{r}
library(mosaic)
library(readr)
library(janitor)
Dexterity <- read_csv("http://nhorton.people.amherst.edu/is5/data/Dexterity.csv") %>%
  clean_names()
```

By default, `read_csv()` prints the variable names. These messages can be suppressed using the `message=FALSE` code chunk option to save space and improve readability.  
Here we use the `clean_names()` function from the `janitor` package to sanitize the names of the columns (which would otherwise contain special characters or whitespace). 

```{r}
Dexterity %>%
  select(age_months, dominant_1, non_dominant_2, gender) %>%
  head(n = 7)
```

### Section 18.1: Paired Data

```{r}
library(tidyr) # for the gather() function
# Figure 18.1, page 586
Dexterity %>%
  select(dominant_1, non_dominant_2) %>%
  gather(key = hand_type, value = speed, dominant_1, non_dominant_2) %>%
  gf_boxplot(speed ~ hand_type, fill = ~ hand_type) %>%
  gf_labs(x = "", y = "Speed (cyl/sec)") +
  ylim(.2, .9) + 
  guides(fill = FALSE)
```

```{r}
Dexterity %>% 
  select(dominant_1, non_dominant_2) %>%
  mutate(difference = dominant_1 - non_dominant_2) %>%
  head(n = 18)
```

### Section 18.2: The Paired *t*-Test

#### Step-By-Step Example: A Paired *t*-Test

XX NH can't find data for 2006 winter olympics speed skating data (page 589)

```{r}
# This data doesn't seem to be correct!! (wrong year an doesn't have lane data)
SpeedySkatey <- read_csv("http://nhorton.people.amherst.edu/is5/data/Winter_Olympics_2010_speed_skating.csv")
SpeedySkatey %>%
  arrange(Result) %>%
  head()
```

### Section 18.3: Confidence Intervals for Matched Pairs

```{r}
Couples <- read_csv("http://nhorton.people.amherst.edu/is5/data/Couples.csv") %>%
  filter(wAge != "*") %>%
  mutate(wAge = as.numeric(wAge))
# table on page 592
Couples %>%
  select(wAge, hAge) %>%
  mutate(difference = hAge - wAge) %>%
  head(n = 7)
```

#### Step-By-Step Example: A Paired *t*-Interval

```{r}
DexData <- Dexterity %>% 
  select(dominant_1, non_dominant_2) %>%
  mutate(difference = dominant_1 - non_dominant_2)
gf_histogram(~ difference, data = DexData, binwidth = .05, center = .025) %>%
  gf_labs(x = "Dominant-Non-dominant", y = "# Subjects")
```

### Section 18.4: Blocking
