---
title: "IS5 in R: Confidence Intervals for Means (Chapter 14)"
author: "Margaret Chien and Nicholas Horton (nhorton@amherst.edu)"
date: "June 27, 2018"
output: 
  pdf_document:
    fig_height: 4
    fig_width: 6
  html_document:
    fig_height: 3
    fig_width: 5
  word_document:
    fig_height: 4
    fig_width: 6
---


```{r, include = FALSE}
# Don't delete this chunk if you are using the mosaic package
# This loads the mosaic and dplyr packages
library(mosaic)
library(readr)
library(janitor)
```

```{r, include = FALSE}
# knitr settings to control how R chunks work.
require(knitr)
opts_chunk$set(
  tidy = FALSE,     # display code as typed
  size = "small"    # slightly smaller font for code
)
```

## Introduction and background 

This document is intended to help describe how to undertake analyses introduced 
as examples in the Fifth Edition of *Intro Stats* (2018) by De Veaux, Velleman, and Bock.
More information about the book can be found at http://wps.aw.com/aw_deveaux_stats_series.  This
file as well as the associated R Markdown reproducible analysis source file used to create it can be found at http://nhorton.people.amherst.edu/is5.

This work leverages initiatives undertaken by Project MOSAIC (http://www.mosaic-web.org), an NSF-funded effort to improve the teaching of statistics, calculus, science and computing in the undergraduate curriculum. In particular, we utilize the `mosaic` package, which was written to simplify the use of R for introductory statistics courses. A short summary of the R needed to teach introductory statistics can be found in the mosaic package vignettes (http://cran.r-project.org/web/packages/mosaic).  A paper describing the mosaic approach was published in the *R Journal*: https://journal.r-project.org/archive/2017/RJ-2017-024.

## Chapter 14: Confidence Intervals for Means

```{r}
library(mosaic)
library(readr)
library(janitor)
Babies <- read_csv("http://nhorton.people.amherst.edu/is5/data/Babysamp_98.csv") %>%
  clean_names()
```

By default, `read_csv()` prints the variable names. These messages can be suppressed using the `message = FALSE` code chunk option to save space and improve readability.  
Here we use the `clean_names()` function from the `janitor` package to sanitize the names of the columns (which would otherwise contain special characters or whitespace).   

```{r, cache=TRUE}
# Figure 14.1, page 441
gf_histogram(~ weight, data = Babies, binwidth = 125) %>%
  gf_labs(x = "Birthweight (g)", y = "# of Babies")

# Figure 14.2
set.seed(12346)
numsim <- 10000
WeightMeans <- do(numsim) * mean(~ weight, data = sample(Babies, 100)) 
gf_histogram(~ mean, data = WeightMeans, binwidth = 10)
favstats(~ mean, data = WeightMeans)
```

### Section 14.1: The Central Limit Theorem

```{r, cache=TRUE}
CEOComp <- read_csv("http://nhorton.people.amherst.edu/is5/data/CEO_Compensation_2014.csv") %>%
  clean_names()
# Figure 14.3
gf_histogram(~ ceo_compensation_m, data = CEOComp, binwidth = 10) %>%
  gf_labs(x = "CEO Compensation in $1,000,000", y = "Frequency")
# Figure 14.4
set.seed(14354)
numsim <- 1000
CEOMeans <- do(numsim) * mean(~ ceo_compensation_m, data = sample(CEOComp, 10))
gf_histogram(~ mean, data = CEOMeans) %>%
  gf_labs(x = "Mean", y = "Frequency")
# Figure 14.5
CEOMeans2 <- do(numsim) * mean(~ ceo_compensation_m, data = sample(CEOComp, 50))
gf_histogram(~ mean, data = CEOMeans2) %>%
  gf_labs(x = "Mean", y = "Frequency")
# Figure 14.7 (skipped 14.6 because it's similar)
CEOMeans3 <- do(numsim) * mean(~ ceo_compensation_m, data = sample(CEOComp, 200))
gf_histogram(~ mean, data = CEOMeans3) %>%
  gf_labs(x = "Mean", y = "Frequency")
```

### Section 14.2: A Confidence Interval for the Mean

```{r}
# Figure 14.9, page 446
gf_dist(dist = "norm", linetype = 2, lwd = 1.5) %>%
  gf_labs(x = "", y = "") %>%
  gf_dist(dist = "t", df = 2, lwd = 1.25) +
  xlim(-4, 4) 
```

#### Example 14.1: A One-Sample *t*-Interval for the Mean

```{r}
Salmon <- read_csv("http://nhorton.people.amherst.edu/is5/data/Farmed_salmon.csv") %>%
  clean_names()
Salmon <- Salmon %>%
  filter(mirex != "NA")
favstats(~ mirex, data = Salmon)
salmonlm <- lm(mirex ~ 1, data = Salmon)
confint(salmonlm, data = Salmon)
```

`confint()` requires the use of an object, such as a linear model.  

#### Example 14.2: Checking Assumptions and Conditions for Student's *t*

```{r}
gf_histogram(~ mirex, data = Salmon, binwidth = 0.04, center = 0.02) %>%
  gf_labs(x = "Mirex (ppm)", y = "# of Salmon")
```

#### Step-By-Step Example: A One-Sample *t*-Interval for the Mean

```{r}
set.seed(34)
BabiesSample <- sample(Babies, 30)
gf_histogram(~ weight, data = BabiesSample, binwidth = 200, center = 100) %>%
  gf_labs(x = "Birthweights", y = "# of Babies")
favstats(~ weight, data = BabiesSample)
babieslm <- lm(weight ~ 1, data = BabiesSample)
confint(babieslm, level = 0.90)
```

### Section 14.3: Interpreting Confidence Intervals

### Section 14.4: Picking Our Interval up by Our Bootstraps

```{r}
gf_histogram(~ mean, data = WeightMeans, binwidth = 10)
```

```{r}
# Figure 14.10, page 454
set.seed(1243536)
numsim <- 10000
bootstrapmeans <- do(numsim) * mean(~ weight, data = mosaic::resample(Babies, 100)) 
gf_histogram(~ mean, data = bootstrapmeans) %>%
  gf_labs(x = "Bootstrap Means", y = "")
# Figure 14.11
bootstraplm <- lm(mean ~ 1, data = bootstrapmeans)
confint(bootstraplm) 
gf_histogram(~ mean, data = bootstrapmeans) %>%
  gf_labs(x = "Bootstrap Means", y = "# of Bootstrap Means")
```

#### Step-By-Step Example: A Bootstrap Confidence Interval for the Mean

```{r}
Commute <- read_csv("http://nhorton.people.amherst.edu/is5/data/Population_Commute_Times.csv") %>%
  clean_names()
set.seed(134)
commutesample <- sample(Commute, 100)
gf_histogram(~ commute_time, data = commutesample, binwidth = 10, center = 5)
# Bootstrap
numsim <- 10000
commutebootstrap <- do(numsim) * mean(~ commute_time, data = mosaic::resample(Commute, 100))
gf_histogram(~ mean, data = commutebootstrap, title = "Time Bootstrap Set Means") %>%
  gf_labs(x = "", y = "")
cbootlm <- lm(mean ~ 1, data = commutebootstrap)
confint(cbootlm)
t.test(commutebootstrap$mean, conf.level = 0.95)
```

### Section 14.5: Thoughts About Confidence Intervals