---
title: "IS5 in R: Stats Starts Here (Chapter 4)"
author: "Margaret Chien and Nicholas Horton (nhorton@amherst.edu)"
date: "June 16, 2018"
output: 
  pdf_document:
    fig_height: 4
    fig_width: 6
  html_document:
    fig_height: 3
    fig_width: 5
  word_document:
    fig_height: 4
    fig_width: 6
---


```{r, include = FALSE}
# Don't delete this chunk if you are using the mosaic package
# This loads the mosaic and dplyr packages
library(mosaic)
library(readr)
library(janitor)
```

```{r, include = FALSE}
# knitr settings to control how R chunks work.
require(knitr)
opts_chunk$set(
  tidy = FALSE,     # display code as typed
  size = "small"    # slightly smaller font for code
)
```

## Introduction and background 

This document is intended to help describe how to undertake analyses introduced 
as examples in the Fifth Edition of *Intro Stats* (2018) by De Veaux, Velleman, and Bock.
More information about the book can be found at http://wps.aw.com/aw_deveaux_stats_series.  This
file as well as the associated R Markdown reproducible analysis source file used to create it can be found at http://nhorton.people.amherst.edu/is5.

This work leverages initiatives undertaken by Project MOSAIC (http://www.mosaic-web.org), an NSF-funded effort to improve the teaching of statistics, calculus, science and computing in the undergraduate curriculum. In particular, we utilize the `mosaic` package, which was written to simplify the use of R for introductory statistics courses. A short summary of the R needed to teach introductory statistics can be found in the mosaic package vignettes (http://cran.r-project.org/web/packages/mosaic).  A paper describing the mosaic approach was published in t
he *R Journal*: https://journal.r-project.org/archive/2017/RJ-2017-024.

## Chapter 4: Understanding and Comparing Distributions

```{r}
library(mosaic)
library(readr)
library(janitor)
HopkinsForest <- read_csv("http://nhorton.people.amherst.edu/is5/data/Hopkins_Forest.csv") %>%
  clean_names()
gf_histogram(~ avg_wind_mph, data = HopkinsForest)
favstats(~ avg_wind_mph, data = HopkinsForest)
```

### Section 4.1: Displays for Comparing Groups

#### Histograms

```{r}
HopkinsForest <- HopkinsForest %>%
  mutate(catmonth = ifelse(month <= 9 & month >= 4, "Spring/Summer", "Fall/Winter"))
gf_histogram(~ avg_wind_mph, data = HopkinsForest) %>%
  gf_facet_wrap(~ catmonth)
favstats(~ avg_wind_mph | catmonth, data = HopkinsForest)
```

#### Example 4.1: Comparing Groups with Stem-And-Leaf

```{r}
NestEgg <- read_csv("http://nhorton.people.amherst.edu/is5/data/Nest_Egg_Index.csv") %>%
  clean_names()
with(NestEgg, stem(nest_egg_index)) # not sure how to group by region
```

#### Boxplots

```{r}
gf_boxplot(avg_wind_mph ~ as.factor(month), data = HopkinsForest) %>%
  gf_labs(x = "month", y = "average wind speed (mph)")
```

#### Example 4.2: Comparing Groups with Boxplots

```{r}
Coasters <- read_csv("http://nhorton.people.amherst.edu/is5/data/Coasters_2015.csv")
gf_boxplot(Track ~ Speed, data = Coasters) %>%
  gf_refine(coord_flip()) # not sure how to make these not lines
# XX MC to fix
```

#### Step-By-Step Example: Comparing Groups

```{r}
Cups <- read_csv("http://nhorton.people.amherst.edu/is5/data/Cups.csv")
favstats(~ Difference | Container, data = Cups)
gf_boxplot(Container ~ Difference, data = Cups) %>%
  gf_refine(coord_flip())
```

#### Just Checking

```{r, warning=FALSE}
Flights <- read_csv("http://nhorton.people.amherst.edu/is5/data/Flights_on_time_2016.csv") %>%
  clean_names()
gf_histogram(~ ontime_pct, data = Flights, binwidth = 2, ylab = "# of Months", xlab = "Ontime %")
gf_boxplot(ontime_pct ~ 1, data = Flights, xlab = "")
gf_boxplot(month ~ ontime_pct, data = Flights)
```

### Section 4.2: Outliers
### Section 4.3: Re-Expressing Data: A First Look

