---
title: "IS5 in R: Inferences for Regression (Chapter 20)"
author: "Margaret Chien and Nicholas Horton (nhorton@amherst.edu)"
date: "July 11, 2018"
output: 
  pdf_document:
    fig_height: 3
    fig_width: 5
  html_document:
    fig_height: 3
    fig_width: 5
  word_document:
    fig_height: 4
    fig_width: 6
---


```{r, include = FALSE}
# Don't delete this chunk if you are using the mosaic package
# This loads the mosaic and dplyr packages
library(mosaic)
library(readr)
library(janitor)
```

```{r, include = FALSE}
# knitr settings to control how R chunks work.
require(knitr)
opts_chunk$set(
  tidy = FALSE,     # display code as typed
  size = "small"    # slightly smaller font for code
)
```

## Introduction and background 

This document is intended to help describe how to undertake analyses introduced 
as examples in the Fifth Edition of *Intro Stats* (2018) by De Veaux, Velleman, and Bock.
More information about the book can be found at http://wps.aw.com/aw_deveaux_stats_series.  This
file as well as the associated R Markdown reproducible analysis source file used to create it can be found at http://nhorton.people.amherst.edu/is5.

This work leverages initiatives undertaken by Project MOSAIC (http://www.mosaic-web.org), an NSF-funded effort to improve the teaching of statistics, calculus, science and computing in the undergraduate curriculum. In particular, we utilize the `mosaic` package, which was written to simplify the use of R for introductory statistics courses. A short summary of the R needed to teach introductory statistics can be found in the mosaic package vignettes (http://cran.r-project.org/web/packages/mosaic).  A paper describing the mosaic approach was published in the *R Journal*: https://journal.r-project.org/archive/2017/RJ-2017-024.

## Chapter 20: Inferences for Regression

```{r}
library(mosaic)
library(readr)
library(janitor)
BodyFat <- read_csv("http://nhorton.people.amherst.edu/is5/data/Bodyfat.csv") %>%
  clean_names()
```

By default, `read_csv()` prints the variable names. These messages can be suppressed using the `message=FALSE` code chunk option to save space and improve readability.  
Here we use the `clean_names()` function from the `janitor` package to sanitize the names of the columns (which would otherwise contain special characters or whitespace).  

```{r}
# Figure 20.1, page 642
gf_point(pct_bf ~ waist, data = BodyFat) %>%
  gf_smooth() %>% # to show linear relationship
  gf_labs(x = "Waist (in.)", y = "% Body Fat")
```

### Section 20.1: The Regression Model

```{r}
lm(pct_bf ~ waist, data = BodyFat)
```

```{r}
# Figure 20.2
gf_histogram(~ pct_bf, data = BodyFat, binwidth =  2.5, center = 1.25) %>%
  gf_labs(x = "% Body Fat", y = "# of Men")
```

XX NH not sure if I can make Figure 20.3 (page 643)

#### Random Matters: Slopes Vary

```{r}
numsamp <- 1000
slopesdata <- do(numsamp) * lm(pct_bf ~ waist, data = resample(BodyFat))
gf_point(pct_bf ~ waist, data = BodyFat) #%>%
#  gf_abline(slope = ~ waist, intercept = ~ Intercept, slopesdata) # gf_coefline()?
gf_histogram(~ waist, data = slopesdata, binwidth = .05, center = .025) %>%
  gf_labs(x = "Slope", y = "# of Trials")
```

### Section 20.2: Assumptions and Conditions

```{r}
# Figure 20.6 is the same as Figure 20.1
# Figure 20.7 (page 645)
bodyfatlm <- lm(pct_bf ~ waist, data = BodyFat)
gf_point(resid(bodyfatlm) ~ fitted(bodyfatlm)) %>%
  gf_labs(x = "Predicted", y = "Residuals")
```

```{r}
Diamonds <- read_csv("http://nhorton.people.amherst.edu/is5/data/Diamonds.csv") %>%
  clean_names()
diamondlm <- lm(price ~ carat_size, data = Diamonds) # there's no carat weight varible?
# Figure 20.8, page 646
gf_point(resid(diamondlm) ~ fitted(diamondlm))
# Figure 20.9
gf_histogram(~ resid(diamondlm))
```

### Section 20.3: Regression Inference and Intuition

### Section 20.4: The Regression Table

### Section 20.5: Multiple Regression Inference

### Section 20.6: Confidence and Prediction Intervals

### Section 20.7: Logistic Regression

### Section 20.8: More About Regression