---
title: "IS5 in R: Scatterplots, Association, and Correlation (Chapter 6)"
author: "Margaret Chien and Nicholas Horton (nhorton@amherst.edu)"
date: "June 16, 2018"
output: 
  pdf_document:
    fig_height: 4
    fig_width: 6
  html_document:
    fig_height: 3
    fig_width: 5
  word_document:
    fig_height: 4
    fig_width: 6
---


```{r, include = FALSE}
# Don't delete this chunk if you are using the mosaic package
# This loads the mosaic and dplyr packages
library(mosaic)
library(readr)
library(janitor)
```

```{r, include = FALSE}
# knitr settings to control how R chunks work.
require(knitr)
opts_chunk$set(
  tidy = FALSE,     # display code as typed
  size = "small"    # slightly smaller font for code
)
```

## Introduction and background 

This document is intended to help describe how to undertake analyses introduced 
as examples in the Fifth Edition of *Intro Stats* (2018) by De Veaux, Velleman, and Bock.
More information about the book can be found at http://wps.aw.com/aw_deveaux_stats_series.  This
file as well as the associated R Markdown reproducible analysis source file used to create it can be found at http://nhorton.people.amherst.edu/is5.

This work leverages initiatives undertaken by Project MOSAIC (http://www.mosaic-web.org), an NSF-funded effort to improve the teaching of statistics, calculus, science and computing in the undergraduate curriculum. In particular, we utilize the `mosaic` package, which was written to simplify the use of R for introductory statistics courses. A short summary of the R needed to teach introductory statistics can be found in the mosaic package vignettes (http://cran.r-project.org/web/packages/mosaic).  A paper describing the mosaic approach was published in t
he *R Journal*: https://journal.r-project.org/archive/2017/RJ-2017-024.

## Chapter 6: Scatterplots, Association, and Correlation

```{r}
library(mosaic)
library(readr)
library(janitor)
Hurricanes <- read_csv("http://nhorton.people.amherst.edu/is5/data/Tracking_hurricanes_2015.csv")
gf_point(Error_72h ~ Year, data = Hurricanes, ylab = "Prediction Error")
```

### Section 6.1: Scatterplots

See dots on pages 164-165.

#### Example 6.1: Comparing Prices Worldwide

```{r}
Prices <- read_csv("http://nhorton.people.amherst.edu/is5/data/Prices_and_Earnings.csv") %>%
  clean_names()
gf_point(food_costs ~ womens_clothing, data = Prices, xlab = "Cost of Women's Clothes", ylab = "Food Costs ($)")
gf_point(i_phone_4s_hr ~ wage_gross, data = Prices, xlab = "Average Hourly Wage", ylab = "Hours to Earn an iPhone 4S")
gf_point(clothing_index ~ hours_worked, data = Prices, xlab = "Working Hours", ylab = "Clothes Index")
gf_point(food_costs ~ vacation_days, data = Prices, xlab = "Vacation Days (per year)", ylab = "Food Costs ($)")
```

#### Roles for Variables

#### Smoothing Scatterplots

```{r}
HopkinsForest <- read_csv("http://nhorton.people.amherst.edu/is5/data/Hopkins_Forest.csv") %>%
  clean_names()
gf_point(avg_wind_mph ~ day_of_year, data = HopkinsForest) %>%
  gf_smooth(se = FALSE, xlab = "Day of Year", ylab = "Average Wind Speed (mph)")
```

#### Example 6.2: Smoothing Timeplots

```{r, message=FALSE}
Fitness <- read_csv("http://nhorton.people.amherst.edu/is5/data/Fitness_data.csv") %>%
  clean_names()
gf_histogram(~ weight, data = Fitness, binwidth = 1, ylab = "# of Days", xlab = "Weight (lb)")
gf_point(weight ~ days_since_july_19_2014, data = Fitness) %>%
  gf_smooth(se = FALSE, xlab = "Days Since July 19, 2014", ylab = "Weight (lb)")
gf_boxplot(weight ~ as.factor(month), data = Fitness, xlab = "Month", ylab = "Weight (lb)")
```

### Section 6.2: Correlation

### Section 6.3: Warning: Correlation $\neq$ Causation

### Section 6.4: Straightening Scatterplots