---
title: "IS5 in R: Comparing Groups (Chapter 17)"
author: "Margaret Chien and Nicholas Horton (nhorton@amherst.edu)"
date: "July 11, 2018"
output: 
  pdf_document:
    fig_height: 3
    fig_width: 5
  html_document:
    fig_height: 3
    fig_width: 5
  word_document:
    fig_height: 4
    fig_width: 6
---


```{r, include = FALSE}
# Don't delete this chunk if you are using the mosaic package
# This loads the mosaic and dplyr packages
library(mosaic)
library(readr)
library(janitor)
```

```{r, include = FALSE}
# knitr settings to control how R chunks work.
require(knitr)
opts_chunk$set(
  tidy = FALSE,     # display code as typed
  size = "small"    # slightly smaller font for code
)
```

## Introduction and background 

This document is intended to help describe how to undertake analyses introduced 
as examples in the Fifth Edition of *Intro Stats* (2018) by De Veaux, Velleman, and Bock.
More information about the book can be found at http://wps.aw.com/aw_deveaux_stats_series.  This
file as well as the associated R Markdown reproducible analysis source file used to create it can be found at http://nhorton.people.amherst.edu/is5.

This work leverages initiatives undertaken by Project MOSAIC (http://www.mosaic-web.org), an NSF-funded effort to improve the teaching of statistics, calculus, science and computing in the undergraduate curriculum. In particular, we utilize the `mosaic` package, which was written to simplify the use of R for introductory statistics courses. A short summary of the R needed to teach introductory statistics can be found in the mosaic package vignettes (http://cran.r-project.org/web/packages/mosaic).  A paper describing the mosaic approach was published in the *R Journal*: https://journal.r-project.org/archive/2017/RJ-2017-024.

## Chapter 17: Comparing Groups

```{r}
library(mosaic)
library(readr)
library(janitor)
```

### Section 17.1: A Confidence Interval for the Difference Between Two Proportions

XX NH need seatbelts data and online profiles data

### Section 17.2: Assumptions and Conditions for Comparing Proportions

### Section 17.3: The Two-Sample *z*-Test: Testing for the Difference Between Proportions

XX NH need sleep data

### Section 17.4: A Confidence Interval for the Difference Between Two Means

### Section 17.5: The Two-Sample *t*-Test: Testing for the Difference Between Two Means

#### Step-By-Step Example: A Two-Sample *t*-Test for the Difference Between the Two Means

```{r}
# page 556
BuyingCam <- read_csv("http://nhorton.people.amherst.edu/is5/data/Buy_from_a_friend.csv")
```

By default, `read_csv()` prints the variable names. These messages can be suppressed using the `message=FALSE` code chunk option to save space and improve readability. 

```{r}
library(tidyr) # for gather() function
BuyingCam <- BuyingCam %>%
  gather(key = buying_type, value = amount_offered, Friend, Stranger)
# Model
gf_boxplot(amount_offered ~ buying_type, fill = ~ buying_type, data = BuyingCam) %>%
  gf_labs(x = "Buying From", y = "Amount Offered ($)", fill = "")
gf_histogram(~ amount_offered, binwidth = 25, center = 12.5, data = BuyingCam) %>% # doesn't exactly match
  gf_facet_wrap(buying_type ~ .) %>%
  gf_labs(x = "Amount Offered", y = "")
# Mechanics
favstats(~ amount_offered | buying_type, data = BuyingCam)
```

### Section 17.6: Randomization Tests and Confidence Intervals for Two Means

### Section 17.7: Pooling

### Section 17.8: The Standard Deviation of a Difference
