---
title: "IS5 in R: Scatterplots, Association, and Correlation (Chapter 6)"
author: "Margaret Chien and Nicholas Horton (nhorton@amherst.edu)"
date: "July 12, 2018"
output: 
  pdf_document:
    fig_height: 3
    fig_width: 5
  html_document:
    fig_height: 3
    fig_width: 5
  word_document:
    fig_height: 4
    fig_width: 6
---


```{r, include = FALSE}
# Don't delete this chunk if you are using the mosaic package
# This loads the mosaic and dplyr packages
library(mosaic)
library(readr)
library(janitor)
```

```{r, include = FALSE}
# knitr settings to control how R chunks work.
require(knitr)
opts_chunk$set(
  tidy = FALSE,     # display code as typed
  size = "small"    # slightly smaller font for code
)
```

## Introduction and background 

This document is intended to help describe how to undertake analyses introduced 
as examples in the Fifth Edition of *Intro Stats* (2018) by De Veaux, Velleman, and Bock.
More information about the book can be found at http://wps.aw.com/aw_deveaux_stats_series.  This
file as well as the associated R Markdown reproducible analysis source file used to create it can be found at http://nhorton.people.amherst.edu/is5.

This work leverages initiatives undertaken by Project MOSAIC (http://www.mosaic-web.org), an NSF-funded effort to improve the teaching of statistics, calculus, science and computing in the undergraduate curriculum. In particular, we utilize the `mosaic` package, which was written to simplify the use of R for introductory statistics courses. A short summary of the R needed to teach introductory statistics can be found in the mosaic package vignettes (http://cran.r-project.org/web/packages/mosaic).  A paper describing the mosaic approach was published in the *R Journal*: https://journal.r-project.org/archive/2017/RJ-2017-024.

## Chapter 6: Scatterplots, Association, and Correlation

```{r}
library(mosaic)
library(readr)
library(janitor)
Hurricanes <- read_csv("http://nhorton.people.amherst.edu/is5/data/Tracking_hurricanes_2015.csv")
```
XX Should eval be set to false in this chunk since it repeats the loading from earlier

By default, `read_csv()` prints the variable names. These messages can be suppressed using the `message=FALSE` code chunk option to save space and improve readability.  

```{r}
# Figure 6.1, page 164
gf_point(Error_72h ~ Year, data = Hurricanes, ylab = "Prediction Error")
```


### Section 6.1: Scatterplots

See dots on pages 164-165.

XX Why include? 

#### Example 6.1: Comparing Prices Worldwide

```{r}
Prices <- read_csv("http://nhorton.people.amherst.edu/is5/data/Prices_and_Earnings.csv") %>%
  clean_names()
names(Prices)
```

Here we use the `clean_names()` function from the `janitor` package to sanitize the names of the columns (which would otherwise contain special characters or whitespace).   

```{r}
gf_point(food_costs ~ womens_clothing, data = Prices) %>%
  gf_labs(x = "Cost of Women's Clothes", y = "Food Costs ($)")
gf_point(i_phone_4s_hr ~ wage_gross, data = Prices) %>%
  gf_labs(x = "Average Hourly Wage", y = "Hours to Earn an iPhone 4S")
gf_point(clothing_index ~ hours_worked, data = Prices) %>%
  gf_labs(x = "Working Hours", y = "Clothes Index")
gf_point(food_costs ~ vacation_days, data = Prices) %>%
  gf_labs(x = "Vacation Days (per year)", y = "Food Costs ($)")
```
XX use of ylab= earlier and use of gf_labs here; be consistent

#### Roles for Variables

#### Smoothing Scatterplots

```{r}
HopkinsForest <- read_csv("http://nhorton.people.amherst.edu/is5/data/Hopkins_Forest.csv") %>%
  clean_names()
# Figure 6.2, page 168
gf_point(avg_wind_mph ~ day_of_year, data = HopkinsForest) %>%
  gf_smooth(se = FALSE) %>%
  gf_labs(x = "Day of Year", y = "Average Wind Speed (mph)")
```

#### Example 6.2: Smoothing Timeplots

```{r, message=FALSE}
Fitness <- read_csv("http://nhorton.people.amherst.edu/is5/data/Fitness_data.csv") %>%
  clean_names()
gf_histogram(~ weight, data = Fitness, binwidth = 1, center = .5) %>%
  gf_labs(x = "Weight (lb)", y = "# of Days")
gf_point(weight ~ days_since_july_19_2014, data = Fitness) %>%
  gf_smooth(se = FALSE) %>%
  gf_labs(x = "Days Since July 19, 2014", y = "Weight (lb)")
gf_boxplot(weight ~ as.factor(month), data = Fitness) %>%
  gf_labs(x = "Month", y = "Weight (lb)")
```

Warnings can be suppressed with the `warnings=FALSE` chunk option. 

XX histogram doesn't match exactly to the one on page 168

### Section 6.2: Correlation

```{r}
HeightsWeights <- read_csv("http://nhorton.people.amherst.edu/is5/data/Heights_and_weights.csv")

# Figure 6.3, page 170
gf_point(Weight ~ Height, data = HeightsWeights) %>%
  gf_labs(x = "Height (in.)", y = "Weight (lb)")
cor(Weight ~ Height, data = HeightsWeights)
```

See displays on pages 170 - 171.

#### Step-by-Step Example: Looking at Association

```{r}
Framingham <- read_csv("http://nhorton.people.amherst.edu/is5/data/Framingham.csv")
gf_point(SBP ~ DBP, data = Framingham) %>%
  gf_labs(x = "Diastolic BP (mm Hg)", y = "Systolic BP (mm Hg)")
cor(SBP ~ DBP, data = Framingham)
```

#### Random Matters: Correlations Vary

```{r, cache=TRUE}
LiveBirths <- read_csv("http://nhorton.people.amherst.edu/is5/data/Babysamp_98.csv") %>%
  clean_names()
LiveBirths <- LiveBirths %>%
  filter(dad_age != "NA")
set.seed(14513)  # To ensure we get the same values when we run it multiple times
numsim <- 10000 # Number of samples
gf_point(mom_age ~ dad_age, data = sample(LiveBirths, size = 50))
# Graph will look different for different samples
cor(mom_age ~ dad_age, data = LiveBirths)

# What does do() do?
cor(mom_age ~ dad_age, data = sample(LiveBirths, size = 50)) # Correlation of one random sample
cor(mom_age ~ dad_age, data = sample(LiveBirths, size = 50)) # Correlation of another random sample

do(2) * cor(mom_age ~ dad_age, data = sample(LiveBirths, size = 50)) # Finds the correlation twice

# For the visualization, we need 10,000 correlations
LiveCorr <- do(numsim) * cor(mom_age ~ dad_age, data = sample(LiveBirths, size = 50))
```

The `do()` function runs, 10,000 times, the correlation and sampling functions on a random sample of 50.  \

XX Since this is the first use of cache=TRUE, should there be a short explanation/intro for its function?  

```{r}
gf_histogram(~ cor, data = LiveCorr) %>%
  gf_labs(x = "Correlation of Mother's Age and Father's Age in Samples of Size 50",
          y = "# of Samples")
```

XX change binwidith to match textbook page 176 
XX include page number

### Section 6.3: Warning: Correlation $\neq$ Causation

```{r}
Storks <- read_csv("http://nhorton.people.amherst.edu/is5/data/Storks.csv")
# Figure 6.9
gf_point(Population ~ Storks, data = Storks) %>%
  gf_labs(x = "# of Storks", y = "Human Population")
```
XX page number

#### Correlation Tables

```{r}
Companies <- read_csv("http://nhorton.people.amherst.edu/is5/data/Companies.csv") %>%
  clean_names()
Companies %>%
  select(assets, sales, market_value, profits, cash_flow, employees) %>%
  cor()
```
XX is hiding either the upper or lower triangles (reflecting) necessary? If not, ignore comment. 
XX page number 

### Section 6.4: Straightening Scatterplots

```{r}
FStops <- read_csv("http://nhorton.people.amherst.edu/is5/data/F-stops.csv") %>%
  clean_names()
# Figure 6.10, page 179
gf_point(f_stop ~ shutter_speed, data = FStops) %>%
  gf_labs(x = "Shutter Speed (sec)", y = "f/stop")
cor(f_stop ~ shutter_speed, data = FStops)
```

#### The Ladder of Powers

#### f/Stops Again

```{r}
# Figure 6.11, page 181
gf_point(log(f_stop) ~ shutter_speed, data = FStops) %>%
  gf_labs(x = "Shutter Speed (sec)", y = "Log (f/stop)")
# Figure 6.12
gf_point((f_stop)^2 ~ shutter_speed, data = FStops) %>%
  gf_labs(x = "Shutter Speed (sec)", y = "f/stop squareed")
```

See displays in "What Can Go Wrong?" on pages 181-183.
