---
title: "IS5 in R: Linear Regression (Chapter 7)"
author: "Margaret Chien and Nicholas Horton (nhorton@amherst.edu)"
date: "June 16, 2018"
output: 
  pdf_document:
    fig_height: 4
    fig_width: 6
  html_document:
    fig_height: 3
    fig_width: 5
  word_document:
    fig_height: 4
    fig_width: 6
---


```{r, include = FALSE}
# Don't delete this chunk if you are using the mosaic package
# This loads the mosaic and dplyr packages
library(mosaic)
library(readr)
library(janitor)
```

```{r, include = FALSE}
# knitr settings to control how R chunks work.
require(knitr)
opts_chunk$set(
  tidy = FALSE,     # display code as typed
  size = "small"    # slightly smaller font for code
)
```

## Introduction and background 

This document is intended to help describe how to undertake analyses introduced 
as examples in the Fifth Edition of *Intro Stats* (2018) by De Veaux, Velleman, and Bock.
More information about the book can be found at http://wps.aw.com/aw_deveaux_stats_series.  This
file as well as the associated R Markdown reproducible analysis source file used to create it can be found at http://nhorton.people.amherst.edu/is5.

This work leverages initiatives undertaken by Project MOSAIC (http://www.mosaic-web.org), an NSF-funded effort to improve the teaching of statistics, calculus, science and computing in the undergraduate curriculum. In particular, we utilize the `mosaic` package, which was written to simplify the use of R for introductory statistics courses. A short summary of the R needed to teach introductory statistics can be found in the mosaic package vignettes (http://cran.r-project.org/web/packages/mosaic).  A paper describing the mosaic approach was published in t
he *R Journal*: https://journal.r-project.org/archive/2017/RJ-2017-024.

## Chapter 7: Linear Regression

```{r}
library(mosaic)
library(readr)
library(janitor)
BurgerKing <- read_csv("http://nhorton.people.amherst.edu/is5/data/Burger_King_items.csv") %>%
  clean_names()
gf_point(fat_g ~ protein_g, data = BurgerKing, xlab = "Protein (g)", ylab = "Fat (g)")
```

### Section 7.1: Least Squares: The Line of "Best Fit"

### Section 7.2: The Linear Model

### Section 7.3: Finding the Least Squares Line

### Section 7.4: Regression to the Mean

### Section 7.5: Examining the Residuals

### Section 7.6: $R^2$--The Variation Accounted For by the Model

### Section 7.7: Regression Assumptions and Conditions