---
title: "IS5 in R: Confidence Intervals for Means (Chapter 14)"
author: "Margaret Chien and Nicholas Horton (nhorton@amherst.edu)"
date: "June 27, 2018"
output: 
  pdf_document:
    fig_height: 4
    fig_width: 6
  html_document:
    fig_height: 3
    fig_width: 5
  word_document:
    fig_height: 4
    fig_width: 6
---


```{r, include = FALSE}
# Don't delete this chunk if you are using the mosaic package
# This loads the mosaic and dplyr packages
library(mosaic)
library(readr)
library(janitor)
```

```{r, include = FALSE}
# knitr settings to control how R chunks work.
require(knitr)
opts_chunk$set(
  tidy = FALSE,     # display code as typed
  size = "small"    # slightly smaller font for code
)
```

## Introduction and background 

This document is intended to help describe how to undertake analyses introduced 
as examples in the Fifth Edition of *Intro Stats* (2018) by De Veaux, Velleman, and Bock.
More information about the book can be found at http://wps.aw.com/aw_deveaux_stats_series.  This
file as well as the associated R Markdown reproducible analysis source file used to create it can be found at http://nhorton.people.amherst.edu/is5.

This work leverages initiatives undertaken by Project MOSAIC (http://www.mosaic-web.org), an NSF-funded effort to improve the teaching of statistics, calculus, science and computing in the undergraduate curriculum. In particular, we utilize the `mosaic` package, which was written to simplify the use of R for introductory statistics courses. A short summary of the R needed to teach introductory statistics can be found in the mosaic package vignettes (http://cran.r-project.org/web/packages/mosaic).  A paper describing the mosaic approach was published in the *R Journal*: https://journal.r-project.org/archive/2017/RJ-2017-024.

## Chapter 14: Confidence Intervals for Means

```{r}
library(mosaic)
library(readr)
library(janitor)
Babies <- read_csv("http://nhorton.people.amherst.edu/is5/data/Babysamp_98.csv") %>%
  clean_names()
```

By default, `read_csv()` prints the variable names. These messages can be suppressed using the `message = FALSE` code chunk option to save space and improve readability.  

```{r}
# Figure 14.1, page 441
gf_histogram(~ weight, data = Babies, binwidth = 125) %>%
  gf_labs(x = "Birthweight (g)", y = "# of Babies")

# Figure 14.2
set.seed(12346)
WeightMeans <- do(10000) * mean(sample(Babies, 100)$weight) %>%
  data.frame() %>%
  rename(means = ".")
gf_histogram(~ means, data = WeightMeans, binwidth = 10)
favstats(~ means, data = WeightMeans)
```

### Section 14.1: The Central Limit Theorem

### Section 14.2: A Confidence Interval for the Mean

### Section 14.3: Interpreting Confidence Intervals

### Section 14.4: Picking Our Interval up by Out Bootstraps

### Section 14.5: Thoughts About Confidence Intervals