---
title: "IS5 in R: XX (Chapter XX)"
author: "Nicholas Horton (nhorton@amherst.edu)"
date: "2025-01-02"
date-format: iso
format: pdf
editor: source
---

```{r}
#| label: setup
#| include: false
library(mosaic)   
library(tidyverse)
```


## Introduction and background 

This document is intended to help describe how to undertake analyses introduced as examples in the Fifth Edition of *Intro Stats* (2018) by De Veaux, Velleman, and Bock.
This file as well as the associated Quarto reproducible analysis source file used to create it can be found at http://nhorton.people.amherst.edu/is5.

This work leverages initiatives undertaken by Project MOSAIC (http://www.mosaic-web.org), an NSF-funded effort to improve the teaching of statistics, calculus, science and computing in the undergraduate curriculum.
In particular, we utilize the `mosaic` package, which was written to simplify the use of R for introductory statistics courses.
A short summary of the R needed to teach introductory statistics can be found in the mosaic package vignettes (https://cran.r-project.org/web/packages/mosaic).
A paper describing the mosaic approach was published in the *R Journal*: https://journal.r-project.org/archive/2017/RJ-2017-024.

We begin by loading packages that will be required for our analyses.

```{r}
library(mosaic)
library(tidyverse)
```

## Chapter X: XX

